Direction-aware functional class scoring enrichment analysis of Infinium DNA methylation data

### 2024-01-04 manuscript/manuscript.Rmd
adrenoreceptors

Another application for this method is the joint pathway enrichment analysis of studies conducted with different
array systems.
Epigenetic differences between infants conceived naturally and with assisted reproductive technologies have been the subject of studies since 2009 [@Katari2009-ya].
Two of the highest powered studies with publicly available data are Estill 2016 which used the 450K array and Novakovic 2019 study which used
the EPIC array [@Estill2016-jh;@Novakovic2019-jt].
Joint enrichment analysis can uncover pathways that are common or different between these studies and illuminate
the biological differences between conception groups.
For this example we focused on comparing naturally conceived infants to those conceived with in vitro fertilisation with fresh embryo transfer.
The Estill dataset on 450k had 418,833 probes that met the quality criteria (Natural n=43; IVF n=38).
Of these, 149,589 probes showed statistical significant differential methylation (FDR<0.05), of which 28,767 and 120,822 exhibited higher or lower methylation in the IVF group respectively.
The Novakovic dataset on EPIC array had 793,844 probes that met the filtering criteria (n=133).
Of these, 5,562 were statistically significant (FDR<0.05) with 1,348 and 4,214 exhibited higher or lower methylation in the IVF group respectively.
After summarisation to the gene level, the Estill dataset on 450k array describes 19,240 genes, while the
Novakovic dataset on EPIC array describes 22,588 genes.
There were 19,234 genes common to both platforms.
The rank-rank plot indicates a high degree of overlap overall, and interestingly shows a trend of more genes having lower methylation level in the IVF group (**Figure 7A**).
A scatterplot of pathway enrichment scores indicates a moderate degree of agreement between these studies at the level of pathway enrichment scores (*r*=0.35,*p*=2.2e-16) despite these studies being conducted on independent cohorts years apart and analysed with different array systems (**Figure 7B**).
A heatmap shows a high degree of agreement of pathways with higher methylation in the IVF groups,
while pathways with overall lower methylation in the IVF group showed some variability in enrichment scores
between studies (**Figure 7C**).
We were interested in which pathways had consistently lower methylation in the IVF group and identified the Reactome pathway "Adrenoreceptors" which had enrichment scores of -0.42 in the Estill study and -0.56 in the Novakovic study, recording a MANOVA FDR value of 0.005 (**Figure 7D**).
Top ranked genes from this pathway include *FOSL1*, *JUNB*, *ASCL1* and *ID1*.
A loss of methylation of these genes could contribute to disorders of development.
Taken together, this analysis shows that LAM method can be used to compare enrichment of independent studies
conducted with 450k and EPIC arrays.

![**Figure 7. Pathway level DNA methylation differences in natural and IVF conceived infants.**
(A) Contour heatmap showing the similarity in gene methylation score ranks in the Estill (450k) and Novakovic (EPIC) studies.
(B) Mitch pathway enrichment scores (S distances) in Estil and Novakovic studies. Pathways with MANOVA FDR<0.05 are shown in red while others are shaded gray.
(C) Heatmap of 30 pathways with most extreme S distances (a surrogate of effect size). Red indicates higher methylation and blue shows lower methylation.
(D) An example of a pathway identified with this joint enrichment analysis method, "Nuclear kinase and transcription factor activation" shows lower methylation of member genes in both Estill and Novakovic studies.
](../figs/fig7.png)

>>>>

Another application for this method is the joint pathway enrichment analysis of studies conducted with different
array systems.
Epigenetic differences between infants conceived naturally and with assisted reproductive technologies have been the subject of studies since 2009 [@Katari2009-ya].
Two of the highest powered studies with publicly available data are Estill 2016 which used the 450K array and Novakovic 2019 study which used
the EPIC array [@Estill2016-jh;@Novakovic2019-jt].
Joint enrichment analysis can uncover pathways that are common or different between these studies and illuminate
the biological differences between conception groups.
For this example we focused on comparing naturally conceived infants to those conceived with in vitro fertilisation with fresh embryo transfer.
The Estill dataset on 450k had 418,833 probes that met the quality criteria (Natural n=43; IVF n=38).
Of these, 149,589 probes showed statistical significant differential methylation (FDR<0.05), of which 28,767 and 120,822 exhibited higher or lower methylation in the IVF group respectively.
The Novakovic dataset on EPIC array had 793,844 probes that met the filtering criteria (n=133).
Of these, 5,562 were statistically significant (FDR<0.05) with 1,348 and 4,214 exhibited higher or lower methylation in the IVF group respectively.
After summarisation to the gene level, the Estill dataset on 450k array describes 19,240 genes, while the
Novakovic dataset on EPIC array describes 22,588 genes.
There were 19,234 genes common to both platforms.
The rank-rank plot indicates a high degree of overlap overall, and interestingly shows a trend of more genes having lower methylation level in the IVF group (**Figure 7A**).
A scatterplot of pathway enrichment scores indicates a moderate degree of agreement between these studies at the level of pathway enrichment scores (*r*=0.35,*p*=2.2e-16) despite these studies being conducted on independent cohorts years apart and analysed with different array systems (**Figure 7B**).
A heatmap shows a high degree of agreement of pathways with higher methylation in the IVF groups,
while pathways with overall lower methylation in the IVF group showed some variability in enrichment scores
between studies (**Figure 7C**).
We were interested in which pathways had consistently lower methylation in the IVF group and identified the Reactome pathway "Adrenoreceptors" which had enrichment scores of -0.42 in the Estill study and -0.56 in the Novakovic study, recording a MANOVA FDR value of 0.04 (**Figure 7D**).
Top ranked genes from this pathway include *ADRA2C*, *ADRB2*, *ADRA2B* and *ADRA1B*.
A loss of methylation of these genes could contribute to reported elevated blood pressure in people conceived by IVF [@Meister2018-cp].
Taken together, this analysis shows that LAM method can be used to compare enrichment of independent studies
conducted with 450k and EPIC arrays.

![**Figure 7. Pathway level DNA methylation differences in natural and IVF conceived infants.**
(A) Contour heatmap showing the similarity in gene methylation score ranks in the Estill (450k) and Novakovic (EPIC) studies.
(B) Mitch pathway enrichment scores (S distances) in Estil and Novakovic studies. Pathways with MANOVA FDR<0.05 are shown in red while others are shaded gray.
(C) Heatmap of 30 pathways with most extreme S distances (a surrogate of effect size). Red indicates higher methylation and blue shows lower methylation.
(D) An example of a pathway identified with this joint enrichment analysis method, "Adrenoreceptors" shows lower methylation of member genes in both Estill and Novakovic studies.
](../figs/fig7.png)
